### 2018-11-30 scp-journal/article-before-submission/abstract.tex
All files as completed BEFORE submission

In this paper, new contributions for model-based testing using Communicating
Sequential Processes (CSP) are presented. For a finite non-terminating CSP
process representing the reference model, finite test suites for checking the
conformance relations trace and failures refinement are presented, and their
completeness (that is, capability to uncover conformity violations) is
proven. The fault domains for which complete failure detection can be
guaranteed are specified by means of normalised transition graphs
representing the failures semantics of finite-state CSP processes. While
complete test suites for CSP processes have been previously investigated by
several authors, a sufficient condition for their finiteness is presented
here for the first time. Moreover, it is shown that the test suites are
optimal in two aspects:~(a)~the maximal length of test traces cannot be
further reduced, and (b)~the nondeterministic behaviour cannot be tested with
smaller or fewer sets of events, without losing the test suite's completeness
property.

>>>>

In this paper, new contributions for model-based testing using Communicating
Sequential Processes (CSP) are presented. For a finite non-terminating CSP
process representing the reference model, finite test suites for checking the
conformance relations traces and failures refinement are presented, and their
completeness (that is, capability to uncover conformity violations) is
proven. The fault domains for which complete failure detection can be
guaranteed are specified by means of normalised transition graphs
representing the failures semantics of finite-state CSP processes. While
complete test suites for CSP processes have been previously investigated by
several authors, a sufficient condition for their finiteness is presented
here for the first time. Moreover, it is shown that the test suites are
optimal in two aspects:~(a)~the maximal length of test traces cannot be
further reduced, and (b)~the nondeterministic behaviour cannot be tested with
smaller or fewer sets of events, without losing the test suite's completeness
property.

### 2018-11-30 scp-journal/article-before-submission/conclusion.tex
All files as completed BEFORE submission

% ==============================================================================
\subsubsection*{Discussion of Further Reductions of the Test Effort}
As has been shown in Theorem~\ref{th:maxtracelen}, the maximal length $pq$ of 
traces to be tested for either  failures refinement or trace refinement cannot be further reduced.
It is noteworthy, however, that when testing finite state machines for equivalence, considerably shorter traces can be used. From the classical results published in~\cite{chow:wmethod,vasilevskii1973}, for example, it follows that the maximal trace length to be executed is less or equal to $3p -q$, which is considerably smaller than $pq$ for $p,q\ge 3$.
As a consequence, the investigation of complete test suites establishing failures or trace equivalence is of considerable interest and will be discussed in a future paper.

It is also known from FSM
testing that it is not necessary to check {\it all} traces up to  
length $pq$ 
when testing for reduction of finite state machines (this corresponds to trace refinement). Notable complete strategies supporting this fact have been presented,
for example,
in~\cite{hierons_testing_2004,DBLP:conf/forte/DorofeevaEY05,petrenko_testing_2011,simao_reducing_2012}.
From~\cite{Huang2017} it is known that complete FSM testing theories can be
translated to other formalisms, such as Extended Finite State Machines,
Kripke Structures, or CSP, resulting in likewise complete test strategies for
the latter. We intend to study translations of several promising FSM
strategies to CSP in the future. These will effectively reduce the 
upper bound for the number of test executions to be performed, which  has been 
shown to be of order $O\big(\card{\Sigma}^{pq-1}\big)$ for our trace refinement tests 
in Section~\ref{sec:complexity}.  The
bound $\binom{n}{\lfloor n/2\rfloor}$ for the number of sets to be used in probing the SUT for illegal
deadlocks, however, cannot be further reduced, as has been established in
Lemma~\ref{lemma:hseta}.

% ==============================================================================
\subsubsection*{Discussion of Adaptive Test Cases}
The tests suggested
in~\cite{Hennessy:1988:ATP:50497,DBLP:conf/icfem/CavalcantiG07} were \emph{preset} in the
sense that the trace to be executed was pre-defined for each test. As a consequence,
the authors of~\cite{DBLP:conf/icfem/CavalcantiG07} introduced \emph{inconclusive}
as a third test result, applicable to the situations where the intended trace
of the execution was blocked, due to legal, but nondeterministic behaviour of the
SUT. We consider this as a disadvantage, since, when aiming at executing a specific trace $s$
before being able to check the test objective---for example, the absence of deadlocks when offering a hitting set $H$ of the minimal acceptances---it may take several tries until the full trace $s$ is accepted by the SUT. Considering the complete testing assumption described in Section~\ref{sec:finitecompletefails}, it may even take $c^{\#s}$ tries to reach the end of the trace $s$, if the SUT can legally block every $s$-event due to nondeterminism, so that $c$ 
trials are required for each event is accepted.

@@

Another distinction of our failures test cases $U_F(j)$ to the tests specified in~\cite{DBLP:conf/icfem/CavalcantiG07} consists in the fact that the former test both trace refinement and
the $conf$ conformance relation in one go, whereas the latter use separate test suites to 
establish these two correctness conditions. Again, we consider the  
structure of the  test cases 
$U_F(j)$ as advantageous, since, when checking acceptance of a hitting set $H$ 
after a trace $s$ for $c$ times according to the complete testing assumption,  
any acceptance of an illegal event $e\in\Sigma - [P/s]^0$ should also be revealed within these $c$ tries.

@@

% ==============================================================================
\subsubsection*{Discussion of Fault Domains}
As already mentioned, the work in~\cite{DBLP:conf/pts/CavalcantiS17} defines
a fault domain as the set of processes that refine a given CSP process.  In
that context, only testing for traces refinement is considered, and the
complete test suites may not be finite. So, the work presented here goes well
beyond what is achieved there. On the other hand,
\cite{DBLP:conf/pts/CavalcantiS17} presents an algorithm for test generation
that can be easily adapted to consider additional selection and termination
criteria, like, for example, the length of the traces used to construct
tests. It would be possible, for instance, to use the bound indicated here.
Moreover, specifying a fault domain as a CSP process allows us to model
domain-specific knowledge using CSP. For example, if an initialisation
component defined by a process $I$ can be regarded as correct without further
testing, we can use $I; RUN$ as a fault domain, to indicate that any SUT of
interest implements $I$ correctly, but afterwards has a potentially arbitrary
behaviour specified by $RUN$. In addition, the work
in~\cite{DBLP:conf/pts/CavalcantiS17} is not restricted to finite or
non-terminating processes.

% ==============================================================================
\subsubsection*{Implications for CSP Model Checking}
As explained in the previous sections, passing a test is specified by
$$(\epass\then\Stop) \lessdet_F (Q\parallel[\Sigma] U_F(j)) \hide \Sigma$$ for
failures testing. If the SUT $Q$ is not a programmed piece of software or an
integrated hardware or software system, but just another CSP process
specification, it is of course possible to verify the pass criterion using
the FDR4 model checker. For checking the refinement relation $P\lessdet_F Q$,
the pass criterion has to be verified for $j=0,\dots,pq-1$, where $p$ and $q$
indicate the number of nodes in $P$'s transition graph and the maximal number
of nodes in $Q$'s graph, respectively (Theorem~\ref{th:failurestest}). Since
the test cases $U_F(j)$ have such a simple structure, it is an interesting
question for further research whether checking $(\epass\then\Stop) \lessdet_F
(Q\parallel[\Sigma] U_F(j)) \hide \Sigma$ for $j=0,\dots,pq-1$ can be faster
than directly checking $P\lessdet_F Q$, as one would do in the usual approach
with FDR4. This is of particular interest, since the checks could be
parallelised on several CPUs. Alternatively it is interesting to investigate
whether the check\footnote{We are grateful to Bill Roscoe for suggesting this
option.}
\[
(\epass\then\Stop) \lessdet_F (Q\parallel[\Sigma] \Intchoice_{j=0}^{pq-1} U_F(j)) \hide \Sigma
\]
may perform better than the check of $P\lessdet_F Q$, since the former allows for
other optimisations in the model checker than the latter.

For a large and complex CSP implementation process $Q$, it may be too time consuming to
generate its normalised transition graph, so that its number $q$ of nodes is
unknown. In such a case the suggested options may still be used as efficient
bug finders: use  $p$ of reference process $P$ as initial $q$-value and
increment $q$ from there, as long as each increment reveals new errors.

% ==============================================================================
\subsubsection*{Conclusion}

In this paper, we have introduced finite complete testing strategies for
model-based testing against finite, non-terminating CSP reference models. The
strategies are applicable to the conformance relations failures refinement
and trace refinement. The underlying fault domains have been defined as the
sets of CSP processes whose normalised transition graphs do not have more
than a given number of additional nodes, when compared to the transition
graph of the reference process. For these domains, finite complete test
suites are available. We have shown for the strategy to check failures
refinement that the way of probing the SUT for illegal deadlocks used in our
test cases is optimal in the sense that it is not possible to guarantee
exhaustiveness with fewer probes. Moreover, the maximal length of the test traces involved cannot be reduced without losing the test suite's completeness property; this holds both for the trace and the failures refinement case.

>>>>

% ==============================================================================
\paragraph{Further Reductions of the Test Effort} As shown in
Theorem~\ref{th:maxtracelen}, the maximal length $pq$ of traces to be tested
for either  failures refinement or traces refinement cannot be further
reduced. It is noteworthy, however, that when testing FSMs for equivalence,
considerably shorter traces can be used. From the classical results published
in~\cite{chow:wmethod,vasilevskii1973}, for example, it follows that the
maximal trace length to be executed is less or equal to $3p -q$, which is
considerably smaller than $pq$ for $p,q\ge 3$. As a consequence, the
investigation of complete test suites establishing failures or trace
equivalence is of considerable interest and will be discussed in a future
paper.

It is also known from FSM testing that it is not necessary to check {\it all}
traces up to length $pq$ when testing for reduction of FSMs (which
corresponds to trace refinement). Notable complete strategies supporting this
fact have been presented, for example,
in~\cite{hierons_testing_2004,DBLP:conf/forte/DorofeevaEY05,petrenko_testing_2011,simao_reducing_2012}.
From~\cite{Huang2017} it is known that complete FSM testing theories can be
translated to other formalisms, such as Extended Finite State Machines,
Kripke Structures, or CSP, resulting in likewise complete test strategies for
the latter. We intend to study translations of several promising FSM
strategies to CSP. These will effectively reduce the upper bound for the
number of test executions to be performed, which  has been shown to be of the
order $O\big(\card{\Sigma}^{pq-1}\big)$ for our traces-refinement tests in
Section~\ref{sec:complexity}.  The bound $\binom{n}{\lfloor n/2\rfloor}$ for
the number of sets to be used in probing the SUT for illegal deadlocks,
however, cannot be further reduced, as established in
Lemma~\ref{lemma:hseta}.

% ==============================================================================
\paragraph{Adaptive Test Cases} The tests suggested
in~\cite{Hennessy:1988:ATP:50497,DBLP:conf/icfem/CavalcantiG07} were
\emph{preset} in the sense that the trace to be executed was pre-defined for
each test. As a consequence, the authors
of~\cite{DBLP:conf/icfem/CavalcantiG07} introduced \emph{inconclusive} as a
third test result, applicable to the situations where the intended trace of
the execution was blocked, due to legal, but nondeterministic behaviour of
the SUT. We consider this as a disadvantage, since, when aiming at executing
a specific trace $s$ before being able to check the test objective---for
example, the absence of deadlocks when offering a hitting set $H$ of the
minimal acceptances---it may take several tries until the full trace $s$ is
accepted by the SUT. Considering the complete testing assumption described in
Section~\ref{sec:finitecompletefails}, it may even take $c^{\#s}$ tries to
reach the end of the trace $s$, if the SUT can legally block every event of
$s$ due to nondeterminism, so that $c$ trials are required for each event is
accepted.

Those authors, later, in the context of a richer algebra based on CSP, have
considered a framework similar to adaptive testing~\cite{CG15}.  They have,
however, considered only traces refinement, and have not studied complexity.

@@

Another distinction of our failures test cases $U_F(j)$ to the tests
specified in~\cite{DBLP:conf/icfem/CavalcantiG07} consists in the fact that
the former test both traces refinement and the $conf$ conformance relation in
one go, whereas the latter use separate test suites to establish these two
correctness conditions. Again, we consider the structure of the  test cases
$U_F(j)$ as advantageous, since, when checking acceptance of a hitting set
$H$ after a trace $s$ for $c$ times according to the complete testing
assumption, any acceptance of an illegal event $e\in\Sigma - [P/s]^0$ should
also be revealed within these $c$ tries.

@@

% ==============================================================================
\paragraph{Fault Domains} As already mentioned, the work
in~\cite{DBLP:conf/pts/CavalcantiS17} defines a fault domain as the set of
processes that refine a given CSP process.  In that context, only testing for
traces refinement is considered, and the complete test suites may not be
finite. So, the work presented here goes well beyond what is achieved there,
but is restricted to finite and nonterminating reference processes. In
addition, \cite{DBLP:conf/pts/CavalcantiS17} presents an algorithm for test
generation that can be adapted to consider additional selection and
termination criteria, like, for example, the length of the traces used to
construct tests. It would be possible, for instance, to use the bound
indicated here. Moreover, specifying a fault domain as a CSP process allows
us to model domain-specific knowledge using CSP. For example, if an
initialisation component defined by a process $I$ can be regarded as correct
without further testing, we can use $I; RUN$ as a fault domain, to indicate
that any SUT of interest implements $I$ correctly, but afterwards has an
arbitrary behaviour specified by $RUN$.

% ==============================================================================
\paragraph{Implications for CSP Model Checking} As explained in the previous
sections, passing a test is characterised by the failures-refinement check
$(\epass\then\Stop) \lessdet_F (Q\parallel[\Sigma] U_F(j)) \hide \Sigma$ for
failures testing. If the SUT $Q$ is not a programmed piece of software or an
integrated hardware or software system, but just another CSP process
specification, it is of course possible to verify the pass criterion using
the FDR4 model checker. For checking the refinement relation $P\lessdet_F Q$,
the pass criterion has to be verified for $j=0,\dots,pq-1$, where $p$ and $q$
indicate the number of nodes in $P$'s transition graph and the maximal number
of nodes in $Q$'s graph, respectively (Theorem~\ref{th:failurestest}). Since
the test cases $U_F(j)$ have such a simple structure, it is an interesting
question for further research whether checking $(\epass\then\Stop) \lessdet_F
(Q\parallel[\Sigma] U_F(j)) \hide \Sigma$ for $j=0,\dots,pq-1$ can be faster
than directly checking $P\lessdet_F Q$, as one would do in the usual approach
with FDR4. This is of particular interest, since the checks could be
parallelised on several CPUs. Alternatively it is interesting to investigate
whether the check of\footnote{We are grateful to Bill Roscoe for suggesting
this option.}
\[
(\epass\then\Stop) \lessdet_F (Q\parallel[\Sigma] \Intchoice_{j=0}^{pq-1} U_F(j)) \hide \Sigma
\]
may perform better than the check of $P\lessdet_F Q$, since the former allows for
other optimisations in the model checker.

For a large implementation process $Q$, it may be too time consuming to
generate its normalised transition graph, so that its number $q$ of nodes is
unknown. In such a case, our testing approach based on fault domains may
still be used as efficient bug finders:~use the number of nodes of the
normalised transition graph of the reference process as the initial value for
$q$ and increment $q$ from there, as long as each increment reveals new
errors.

% ==============================================================================
\paragraph{Conclusion}
In this paper, we have introduced finite complete testing strategies for
model-based testing against finite, non-terminating CSP reference models. The
strategies are applicable to the conformance relations failures refinement
and traces refinement. The underlying fault domains have been defined as the
sets of CSP processes whose normalised transition graphs do not have more
than a given number of additional nodes, when compared to the transition
graph of the reference process. For these domains, finite complete test
suites are available. We have shown for the strategy to check failures
refinement that the way of probing the SUT for illegal deadlocks in our test
cases is optimal, so that it is not possible to guarantee exhaustiveness with
fewer probes. Moreover, the maximal length of the test traces cannot be
reduced without losing the test suite's completeness property; this holds
both for traces and failures refinement.

### 2018-11-30 scp-journal/article-before-submission/introduction.tex
All files as completed BEFORE submission

In~\cite{Hennessy:1988:ATP:50497,DBLP:conf/icfem/CavalcantiG07,DBLP:conf/pts/CavalcantiS17},
the authors devised linear test cases: after running through a preset trace
$s$, test cases for trace refinement check for illegal acceptance of a
specific event $e$, and test cases verifying  nondeterministic behaviour
check for  acceptance of events from some set $A$. In the present article, we
follow the alternative approach proposed
in~\cite{DBLP:conf/fm/PeleskaS96,peleska1997a} and use {\it adaptive} test
cases. This means that each test case adapts its trace execution to the
nondeterministic behaviour of the SUT, checks for trace violations at any
point during the test execution, and checks for the acceptance of a given
minimal hitting set of events after any legal trace of a test case-specific
length.

The adaptive test cases have the advantage that test executions only lead to
an inconclusive result if the reference process allows for a nondeterministic
choice between deadlock and trace continuation in a certain state. In
contrast to this, the linear test cases may lead to many more futile
executions with inconclusive results, if the SUT refuses to engage into the
next event $e$ from the preset trace $s$, due to legal nondeterministic
choices leading to a refusal of $e$. Moreover, the preset traces $s$ need to
be executed twice according to the strategies devised in
\cite{Hennessy:1988:ATP:50497,DBLP:conf/icfem/CavalcantiG07,DBLP:conf/pts/CavalcantiS17},
because trace refinement and correctness of nondeterministic behaviour are
checked by two different sets of test cases.

@@

\paragraph{Overview} In Section~\ref{section:preliminaries}, we present the
background relevant to our work. In Section~\ref{sec:finitecompletefails},
finite complete test suites for verifying failures refinement are presented.
A sample test suite is presented in Section~\ref{sec:case}. Test suites
checking trace refinement are a simplified version of the former class; they
are presented in Section~\ref{sec:finitecomplete}. The optimality results are
presented in Section~\ref{sec:complexity}, together with further complexity
considerations. Our results are discussed in Section~\ref{sec:conc}, where we
also conclude. References to further related work are given throughout the
paper where appropriate.

>>>>

In~\cite{Hennessy:1988:ATP:50497,DBLP:conf/icfem/CavalcantiG07,DBLP:conf/pts/CavalcantiS17},
the authors devised linear test cases:~after running through a preset trace
$s$, test cases for traces refinement check for illegal acceptance of a
specific event $e$, and test cases verifying  nondeterministic behaviour
check for  acceptance of events from some set $A$. In the present article, we
follow the alternative approach proposed
in~\cite{DBLP:conf/fm/PeleskaS96,peleska1997a} and use {\it adaptive} test
cases. This means that each test case adapts its trace execution to the
nondeterministic behaviour of the SUT, checks for trace violations at any
point during the test execution, and checks for the acceptance of a given
minimal hitting set of events after any legal trace of a test case-specific
length.

The adaptive test cases have the advantage that test executions only lead to
an inconclusive result if the reference process allows for a nondeterministic
choice between deadlock and trace continuation in a certain state. In
contrast to this, the linear test cases may lead to many more futile
executions with inconclusive results, if the SUT refuses to engage into the
next event $e$ from the preset trace $s$, due to legal nondeterministic
choices leading to a refusal of $e$. Moreover, the preset traces $s$ need to
be executed twice according to the strategies devised in
\cite{Hennessy:1988:ATP:50497,DBLP:conf/icfem/CavalcantiG07,DBLP:conf/pts/CavalcantiS17},
because traces refinement and correctness of nondeterministic behaviour are
checked by two different sets of test cases.

@@

\paragraph{Overview} In Section~\ref{section:preliminaries}, we present the
background relevant to our work. In Section~\ref{sec:finitecompletefails},
finite complete test suites for verifying failures refinement are presented.
A sample test suite is presented in Section~\ref{sec:case}. Test suites
checking traces refinement are a simplified version of the former class; they
are presented in Section~\ref{sec:finitecomplete}. The optimality results are
presented in Section~\ref{sec:complexity}, together with further complexity
considerations. Our results are discussed in Section~\ref{sec:conc}, where we
also conclude. References to further related work are given throughout the
paper where appropriate.

### 2018-11-30 scp-journal/article-before-submission/preliminaries.tex
All files as completed BEFORE submission

For divergence-free processes, failures refinement, $P \lessdet_F Q$, is
defined by $\fails(Q)\subseteq\fails(P)$. Since refusals are subset-closed,
$P \lessdet_F Q$ implies $(s,\varnothing)\in\fails(P)$ for all traces
$s\in\trc(Q)$. So, for divergence-free processes, failures refinement implies
trace refinement. Therefore, using the conformance relation $conf$ below
%
\begin{equation}\label{eq:conf}
  Q\ conf\ P \defs \forall s\in \trc(P) \cap \trc(Q): \refs(Q/s)
  \subseteq \refs(P/s),
\end{equation}
%
failures refinement can be expressed by $\lessdet_T$ and $conf$ as proven
in~\cite{DBLP:conf/icfem/CavalcantiG07}.
%
\begin{equation}%\label{thm:fref-tref-conf}
\label{eq:failconf}
(P \lessdet_F Q) \iff (P \lessdet_T Q \land Q\ conf\ P)
\end{equation}
%
For finite processes, since refusals are subset-closed, $\refs(P/s)$ can be
constructed from the set of \emph{maximal refusals}.
%
\begin{equation}
\maxrefs(P/s) = \{ R \in\refs(P/s)~|~\forall R'\in \refs(P/s) - \{ R\}: R \not\subseteq R'\}
\end{equation}
%
Conversely, with the maximal refusals $\maxrefs(P/s)$ at hand, we can
reconstruct the refusals in the set $\refs(P/s)$. % as described by
%
\begin{equation}\label{eq:refMaxRef}
\refs(P/s) = \{ R'\in 2^\Sigma~|~\exists R\in \maxrefs(P/s): R'\subseteq R \}.
\end{equation}
%
Deterministic process states $P/s$ have exactly the one maximal refusal
defined by $\Sigma-[P/s]^0$, where $[P/s]^0$ denotes the \emph{initials} of
$P/s$, that is, the events that $P/s$ may engage into. The maximal refusals
in combination with the initials of a process express its degree of
nondeterminism as illustrated by the next example.
%
\begin{example}
\label{ex:nondetdegree} $P = (\Stop \intchoice Q)$ has maximal refusals
$\maxrefs(P) = \{ \Sigma \}$, because $\Stop$ refuses to engage in any event,
and this is carried over to $P$ by the internal choice. However, $P$ is
distinguished from $\Stop$ by its initials, which are defined by $[P]^0 = [
\Stop \intchoice Q]^0 = [Q]^0$. So $P$ may engage nondeterministically in any
initial event of $Q$, but also refuse everything, due to internal selection
of $\Stop$. Assuming an alphabet $\Sigma = \{a,b,c,d\}$, the process
%
$$Q = (e:\{a,b\} \then \Stop) \intchoice (e:\{c,d\} \then \Stop)$$
%
has maximal refusals $\maxrefs(Q) = \{ \{c,d\}, \{a,b\} \}$ and initials
$[Q]^0=\Sigma$. In contrast to $P$, nondeterminism is reflected here by
two maximal refusals. \xbox
\end{example}

@@

(\ref{eq:trcrefa}) reflects trace refinement in terms of graph traces; result
(\ref{eq:failconf}) expresses failures refinement in terms of traces

>>>>

For divergence-free processes, failures refinement, $P \lessdet_F Q$, is
defined by $\fails(Q)\subseteq\fails(P)$. Since refusals are subset-closed,
$P \lessdet_F Q$ implies $(s,\varnothing)\in\fails(P)$ for all traces
$s\in\trc(Q)$. So, for divergence-free processes, failures refinement implies
traces refinement. Therefore, using the conformance relation $conf$ below
%
\begin{equation}\label{eq:conf}
  Q\ conf\ P \defs \forall s\in \trc(P) \cap \trc(Q): \refs(Q/s)
  \subseteq \refs(P/s),
\end{equation}
%
failures refinement can be expressed by $\lessdet_T$ and $conf$ as proven
in~\cite{DBLP:conf/icfem/CavalcantiG07}.
%
\begin{equation}%\label{thm:fref-tref-conf}
\label{eq:failconf}
(P \lessdet_F Q) \iff (P \lessdet_T Q \land Q\ conf\ P)
\end{equation}
%
For finite processes, since refusals are subset-closed, $\refs(P/s)$ can be
constructed from the set of \emph{maximal refusals}.
%
\begin{equation}
\maxrefs(P/s) = \{ R \in\refs(P/s)~|~\forall R'\in \refs(P/s) - \{ R\}: R \not\subseteq R'\}
\end{equation}
%
Conversely, with the maximal refusals $\maxrefs(P/s)$ at hand, we can
reconstruct the refusals in the set $\refs(P/s)$. % as described by
%
\begin{equation}\label{eq:refMaxRef}
\refs(P/s) = \{ R'\in 2^\Sigma~|~\exists R\in \maxrefs(P/s): R'\subseteq R \}.
\end{equation}
%
Deterministic process states $P/s$ have exactly the one maximal refusal
defined by $\Sigma-[P/s]^0$, where $[P/s]^0$ denotes the \emph{initials} of
$P/s$, that is, the events that $P/s$ may engage into. The maximal refusals
in combination with the initials of a process express its degree of
nondeterminism as illustrated by the next example.
%
\begin{example}
\label{ex:nondetdegree} $P = (\Stop \intchoice Q)$ has maximal refusals
$\maxrefs(P) = \{ \Sigma \}$, because $\Stop$ refuses to engage in any event,
and this is carried over to $P$ by the internal choice. However, $P$ is
distinguished from $\Stop$ by its initials, which are defined by $[P]^0 = [
\Stop \intchoice Q]^0 = [Q]^0$. So $P$ may engage nondeterministically in any
initial event of $Q$, but also refuse everything, due to internal selection
of $\Stop$. Assuming an alphabet $\Sigma = \{a,b,c,d\}$, the process
%
$$Q = (e:\{a,b\} \then \Stop) \intchoice (e:\{c,d\} \then \Stop)$$
%
has maximal refusals $\maxrefs(Q) = \{ \{c,d\}, \{a,b\} \}$ and initials
$[Q]^0=\Sigma$. In contrast to $P$, nondeterminism is reflected here by
two maximal refusals. \xbox
\end{example}

@@

(\ref{eq:trcrefa}) reflects traces refinement in terms of graph traces;
result (\ref{eq:failconf}) expresses failures refinement in terms of traces
